Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-06 aistats/paper.tex
attempt to dump a few optim papers

Then; 
- description of the setting
- statistical solutions (asymptotic, l1 or l2 error)
- optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
- potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
- conclusion?

@@

@@

\section{Insight - Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We can get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where the norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian at the optimum 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1}  \; .
\end{align}
Both the MLE and the MAP have the same asymptotic rate
\begin{align}
\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data.
Remark how this rate is also independent of the optimum $\meanp^*$.

@@

@@

\section{Stuff to cite and why }

\begin{itemize}
	\item Stochastic and online mirror descent: 
	We obviously have rates in this setting, but they all assume 
	that the reference function is strongly-convex relative to a norm. 
	\citep[e.g.][Thm 4.2]{bubeck2015convex}
	\item What is mirror descent called in the literature?
		\begin{itemize}
			\item Mirror Descent \citep{nemirovski1983problem,beck2003mirror} 
			\item NoLips \citep{bauschke2017descent}
		\end{itemize}
	\item Interest in mirror descent boosted by the relatively smooth view 
	of \citet{bauschke2017descent,lu2018relatively}	
	\item 
	The oroblems on how to deal with Bregman divergences abound in optimization, 
	beyond stochasticity. 
	For example, we haven't yet figured out the analog of Nesterov-type acceleration 
	on relatively smooth and strongly convex problems, 
	to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
	or just from $1/T$ to $1/T^2$ 
	in the (non-strongly) convex case.
	\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
	The tools developped to make progress on one problem might help make progress on the other. 
	\item 
	Stochastic mirror descent in the relatively smooth setting;
	\citet{hanzely2018fastest},
	\citet{dragomir2021fast}
	and the zoo of assumptions
	\item Something about the statistical asymptotic behavior for the MLE? 
	\begin{equation}
		D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
		\approx \frac{1}{n}?
	\end{equation}
	\item Something about how the Bregman divergence is often not considered in the statistics lit 
	because it can be infinite and is non-symmetric and not nice to work with, 
	and so folks focus on $\expect{\norm{\theta_n - \theta_*}^2}$ even if it's somewhat meaningless 
	in the non-asymptotic regime, 
	and the best non-asymptotic results 
	still rely on Taylor expansions after $n > $ something big?
	\item Something about how we don't even know how to deal with decreasing step-sizes 
	of the order of $1/t$ 
	\item 
	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions 
	\url{https://www.di.ens.fr/~fbach/bach_ejs_self_concordance.pdf}
	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
	which is the ``expected'' rate for that setting, 
	and how this approach above is a generalization beyond the Euclidean noise model 
	\url{http://proceedings.mlr.press/v99/marteau-ferey19a/marteau-ferey19a.pdf}
	\url{http://proceedings.mlr.press/v99/ostrovskii19a/ostrovskii19a.pdf}
\end{itemize}

>>>>

% Then; 
% - description of the setting
% - statistical solutions (asymptotic, l1 or l2 error)
% - optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
% - potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
% - conclusion?

@@

\fdk{
Might want to note that a perhaps ``typical'' view of parameter estimation 
works with norms, e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ (for some norm and some power), 
or more generally using distances rather than divergences as they are symmetric and have nicer properties. 
But as the MLE is the minimizer of the log-likelihood, 
using the log-likelihood (and by extension the KL divergence) as a measure of risk is natural, if inconvenient.
}

@@

\section{Insight - Asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We can get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
    + O(\norm{\mu - \mu^*}^3)
\end{align}
where the norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian at the optimum 
\begin{align}
    \nabla^2\conj(\mu^*) 
    = \nabla^2\logpart(\natp^*)^{-1} 
    = \Cov_{\natp^*}[T(X)]^{-1}  \; .
\end{align}
Both the MLE and the MAP have the same asymptotic rate
\begin{align}
\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
\end{align}
as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data.
Remark how this rate is also independent of the optimum $\meanp^*$.
\fdk{meaning "independent ignoring constants"?}

\fdk{
That should be well known, maybe find a reference?
% Something about the statistical asymptotic behavior for the MLE? 
% \begin{equation}
% 	D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
% 	\approx \frac{1}{n}?
% \end{equation}
}

@@

\section{Optimization related work}
In the context of optimization, why do we care?

There is a long line work trying the convergence rate of stochastic gradient descent.
Beyond the results of RM51, 
we now have proofs n smooth, strongly convex problems and bounded variance 
(as well as other assumptions on the noise, reviewed later)
a decreasing step-size and an averaging scheme gets 
a $1/t$ convergence rate, 
which matches the asymptotic rate of unbiased estimation,
through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler}\footnote{
Those papers assume the stochastic gradients are bounded, 
but small modifications also work for bounded variance instead.
}

We do have extensions to other notions of bounded variance, 
for example assuming that the stochastic gradients are gradients of a perturbed smooth function $f_i$ 
and that the minimum of $f$ and the minima of the $f_i$ 
are bounded, $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$,
or that the gradient noise is bounded only at the minimum. 
For a review, see \citet{gower2019sgd}.

However, for some problems, combining decreasing step-sizes and averaging is not necessary,
even when the function is not strongly convex.
This is the case for maximum likelihood estimation and matches the asymptotic rates, 
but holds more generally, for example for linear and logistic regressions \citep{bach2013non}.

There has been recent advances in optimization that go beyond the Euclidean case, 
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
\citep{bauschke2017descent,lu2018relatively},
where instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
we use a reference function instead \fdk{maybe add a line}
and apply mirror descent \citep{nemirovski1983problem,beck2003mirror}
(also called NoLips \citep{bauschke2017descent}, something else?).

We have convergence results for the 
application of stochastic mirror descent, 
but they typically rely on assumptions that are too strong. 
\begin{itemize}
	\item The reference function is strongly convex [typical setup in online learning]
	\item The variance is uniformly bounded
\end{itemize}
Some advances have been tried
\begin{itemize}
	\item The reference function is ``close to'' a norm
	and looks like a \citet{hanzely2018fastest}
	\item 
	Attempt to adapt the ``bounded gradient at the minimum''
	\citep{dragomir2021fast}
\end{itemize}

Other to cite;
\begin{itemize}
	\item \citet{marteauferey2019}
\end{itemize}

\fdk{
A ``simple open problem'';
assume we have a deterministic problem, so that we know where the minimum is. 
Can we figure out a (deterministic) path from $\theta_0$ to $\theta_*$ 
that has optimality decreasing as $1/n^2$, 
to mirror the bias term of gradient descent?
}

\fdk{
Application

The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
For example, we haven't yet figured out the analog of Nesterov-type acceleration 
on relatively smooth and strongly convex problems, 
to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
or just from $1/T$ to $1/T^2$ 
in the (non-strongly) convex case.
\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
The tools developped to make progress on one problem might help make progress on the other. 
}

@@
